The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-06 script/en/chapters/cryptomethods.tex
resolved more pdfTeX warnings by adding a chapter specific prefix "xy:" to all cite and bibitem in chapters with bibliography

The book of Bruce Schneier \cite{Schneier1996cm} offers a very good overview of
the different algorithms. We basically distinguish between symmetric and
asymmetric encryption procedures.

@@

All classical methods are of this type. Examples can be found within the 
program CrypTool, in chapter  \ref{Kapitel_PaperandPencil} 
(Paper and Pencil Encryption Methods) of this script or in \cite{Nichols1996}. 
Now we want to consider more modern mechanisms.

@@

Below you will find some results, which have recently called into question the security of the AES algorithm -- from our point of view these doubts practically still remain unfounded
% = do not bring disrepute upon AES
. 
The following information is based on the original papers and the articles \cite{Wobst-iX2002} and \cite{Lucks-DuD2002}.

@@

In their 2001-paper Ferguson, Schroeppel and Whiting \cite{Ferguson2001}
presented a new method of symmetric codes cryptanalysis: They described AES with
a closed formula (in the form of a continued fraction) which was possible
because of the "relatively" clear structure of AES. This formula consists of
around 1000 trillion terms of a sum - so it does not help concrete practical
cryptanalysis. Nevertheless curiosity in the academic community was awakened.
It was already known, that the 128-bit AES could be described as an
over-determined system of about 8000 quadratic equations (over an algebraic
number field) with about 1600 variables (some of them are the bits of the wanted
key) -- equation systems of that size are in practice not solvable. This special
equation system is relatively sparse, so only very few of the quadratic terms
(there are about 1,280,000 are possible quadratic terms in total) appear in the
equation system.

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product algorithm\index{Cascade cipher}%
\footnote{%
A cipher text can be used as input for another encryption algorithm. 
A cascade cipher is build up as a composition of different encryption 
transformations. The overall cipher is called product algorithm or cascade cipher (sometimes depending whether the used keys are statistically dependent or not).\\
Cascading does not always improve the security.\\
This process is also used {\em within} modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively insecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher. 
}
does not exponentially increase with the number of rounds.

Currently there is a large amount of research in this area: for example Murphy and Robshaw presented a paper at Crypto 2002 \cite{Robshaw2002a}, which could dramatically improve cryptanalysis: the burden for a 128-bit key was estimated at about $2^{100}$ steps by describing AES as a special case of an algorithm called BES (Big Encryption System), which has an especially "round" structure. But even $2^{100}$ steps are beyond what is achievable in the foreseeable future. Using a 256 bit key the authors estimate that a XSL-attack will require $2^{200}$ operations.

@@

The discussion is very controversial at the moment: Don Coppersmith (one of the
inventors of DES) for example queries the practicability of the attack because
XLS would provide no solution for AES \cite{Coppersmith2002}. This implies that
then the optimization of Murphy and Robshaw \cite{Robshaw2002b} would not work.

In 2009 Biryukov und Khovratovich \cite{Biryukov2009} published another
theoretical attack on AES. This attack uses different methods from the ones
described above. They applied methods from hash function cryptanalysis (local
collisions and boomerang switching) to construct a related-key attack on
AES-256. I.~e.\ the attacker not only needs to be able to encrypt arbitrary data
(chosen plain text), in addition he needs to be able to manipulate the unknown key
(related-key).

@@

If Alice\index{Alice}%
\footnote{%
      In order to describe cryptographic protocols participants
      are often named Alice, Bob\index{Bob}, \dots (see \cite[p. 23]{Schneier1996cm}). 
      Alice and Bob perform all 
      2-person-protocols. Alice will initiate all protocols and 
      Bob answers. The attackers are named Eve (eavesdropper) and
      Mallory (malicious active attacker).
  } wants to communicate with Bob, then she finds Bob's public key 
in the directory and uses it to encrypt her message to him. She then sends
this cipher text to Bob, who is then able to decrypt it again using his 
secret key. As only Bob knows his secret key, only he can decrypt 
messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

@@

\bibitem[Biryukov2009]{Biryukov2009} \index{Biryukov 2009}
	Alex Biryukov, Dmitry Khovratovich, \\
	{\em Related-key Cryptanalysis of the Full AES-192 and AES-256}, \\
	2009 \\
	\url{http://eprint.iacr.org/2009/317}

\bibitem[Coppersmith2002]{Coppersmith2002}  \index{Coppersmith 2002}
       Don Coppersmith, \\
       {\em Re: Impact of Courtois and Pieprzyk results}, \\
       2002-09-19, ``AES Discussion Groups''~ at \\
       \href{http://aes.nist.gov/aes/}
       {\texttt{http://aes.nist.gov/aes/}}

\bibitem[Courtois2002]{Courtois2002}  \index{Courtois 2002}
        Nicolas Courtois, Josef Pieprzyk, \\
        {\em Cryptanalysis of Block Ciphers with Overdefined Systems
             of Equations}, \\
        received 10 Apr 2002, last revised 9 Nov 2002.\\
        A different version, so called compact version of the first XSL attack,
       was published at Asiacrypt Dec 2002. \\
       \href{http://eprint.iacr.org/2002/044}
       {\texttt{http://eprint.iacr.org/2002/044}}

\bibitem[Ferguson2001]{Ferguson2001}  \index{Ferguson 2001}
       Niels Ferguson, Richard Schroeppel, Doug Whiting, \\
       {\em A simple algebraic representation of Rijndael}, 
       Draft 2001/05/1, \\
       \href{http://www.xs4all.nl/~vorpal/pubs/rdalgeq.html}
       {\texttt{http://www.xs4all.nl/\~{}vorpal/pubs/rdalgeq.html}}

\bibitem[Lucks-DuD2002]{Lucks-DuD2002}  \index{Lucks 2002}
       Stefan Lucks, R\"udiger Weis, \\
       {\em Neue Ergebnisse zur Sicherheit des Verschl\"usselungsstandards AES}, 
       in DuD Dec. 2002.

\bibitem[Nichols1996]{Nichols1996} \index{Nichols 1996} 
       Randall K. Nichols, \\
       {\em Classical Cryptography Course, Volume 1 and 2}, \\
       Aegean Park Press 1996;
       or in 12 lessons online at \\
       \href{http://www.fortunecity.com/skyscraper/coding/379/lesson1.htm}
       {\texttt{http://www.fortunecity.com/skyscraper/coding/379/lesson1.htm}}

\bibitem[Robshaw2002a]{Robshaw2002a}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Essential Algebraic Structure within the AES}, 
       June 5, 2002, Crypto 2002,  \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

\bibitem[Robshaw2002b]{Robshaw2002b}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Comments on the Security of the AES and the XSL Technique}, 
       September 26, 2002, \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

\bibitem[Schmeh2003]{Schmeh2003}  \index{Schmeh 2003}
       Klaus Schmeh, \\
       {\em Cryptography and Public Key Infrastructures on the Internet},\\ 
       John Wiley \& Sons Ltd., Chichester 2003. \\
       An up-to-date, easy to read book, which also considers practical
       problems such as standardisation or real existing software.

\bibitem[Schneier1996]{Schneier1996cm} \index{Schneier 1996} 
       Bruce Schneier, \\
       {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
       Wiley 1994, 2nd edition 1996.

\bibitem[Stamp2007]{Stamp2007} \index{Stamp 2007} 
       Mark Stamp, Richard M. Low, \\
       {\em Applied Cryptanalysis: Breaking Ciphers in the Real World}, \\
       Wiley-IEEE Press 2007, \\
       \href{http://cs.sjsu.edu/faculty/stamp/crypto/}
       {\texttt{http://cs.sjsu.edu/faculty/stamp/crypto/}}

\bibitem[Swenson2008]{Swenson2008} \index{Swenson 2008} 
       Christopher Swenson, \\
       {\em Modern Cryptanalysis: Techniques for Advanced Code Breaking}, \\
       Wiley 2008.

\bibitem[Wobst-iX2002]{Wobst-iX2002}  \index{Wobst 2002}
       Reinhard Wobst, \\
       {\em Angekratzt - Kryptoanalyse von AES schreitet voran}, 
       in iX Dec. 2002, \\
       plus the reader's remark by Johannes Merkle in iX Feb. 2003.

>>>>

The book of Bruce Schneier \cite{cm:Schneier1996cm} offers a very good overview of
the different algorithms. We basically distinguish between symmetric and
asymmetric encryption procedures.

@@

All classical methods are of this type. Examples can be found within the 
program CrypTool, in chapter  \ref{Kapitel_PaperandPencil} 
(Paper and Pencil Encryption Methods) of this script or in \cite{cm:Nichols1996}. 
Now we want to consider more modern mechanisms.

@@

Below you will find some results, which have recently called into question the security of the AES algorithm -- from our point of view these doubts practically still remain unfounded
% = do not bring disrepute upon AES
. 
The following information is based on the original papers and the articles \cite{cm:Wobst-iX2002} and \cite{cm:Lucks-DuD2002}.

@@

In their 2001-paper Ferguson, Schroeppel and Whiting \cite{cm:Ferguson2001}
presented a new method of symmetric codes cryptanalysis: They described AES with
a closed formula (in the form of a continued fraction) which was possible
because of the "relatively" clear structure of AES. This formula consists of
around 1000 trillion terms of a sum - so it does not help concrete practical
cryptanalysis. Nevertheless curiosity in the academic community was awakened.
It was already known, that the 128-bit AES could be described as an
over-determined system of about 8000 quadratic equations (over an algebraic
number field) with about 1600 variables (some of them are the bits of the wanted
key) -- equation systems of that size are in practice not solvable. This special
equation system is relatively sparse, so only very few of the quadratic terms
(there are about 1,280,000 are possible quadratic terms in total) appear in the
equation system.

The mathematicians Courtois and Pieprzyk \cite{cm:Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product algorithm\index{Cascade cipher}%
\footnote{%
A cipher text can be used as input for another encryption algorithm. 
A cascade cipher is build up as a composition of different encryption 
transformations. The overall cipher is called product algorithm or cascade cipher (sometimes depending whether the used keys are statistically dependent or not).\\
Cascading does not always improve the security.\\
This process is also used {\em within} modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively insecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher. 
}
does not exponentially increase with the number of rounds.

Currently there is a large amount of research in this area: for example Murphy and Robshaw presented a paper at Crypto 2002 \cite{cm:Robshaw2002a}, which could dramatically improve cryptanalysis: the burden for a 128-bit key was estimated at about $2^{100}$ steps by describing AES as a special case of an algorithm called BES (Big Encryption System), which has an especially "round" structure. But even $2^{100}$ steps are beyond what is achievable in the foreseeable future. Using a 256 bit key the authors estimate that a XSL-attack will require $2^{200}$ operations.

@@

The discussion is very controversial at the moment: Don Coppersmith (one of the
inventors of DES) for example queries the practicability of the attack because
XLS would provide no solution for AES \cite{cm:Coppersmith2002}. This implies that
then the optimization of Murphy and Robshaw \cite{cm:Robshaw2002b} would not work.

In 2009 Biryukov und Khovratovich \cite{cm:Biryukov2009} published another
theoretical attack on AES. This attack uses different methods from the ones
described above. They applied methods from hash function cryptanalysis (local
collisions and boomerang switching) to construct a related-key attack on
AES-256. I.~e.\ the attacker not only needs to be able to encrypt arbitrary data
(chosen plain text), in addition he needs to be able to manipulate the unknown key
(related-key).

@@

If Alice\index{Alice}%
\footnote{%
      In order to describe cryptographic protocols participants
      are often named Alice, Bob\index{Bob}, \dots (see \cite[p. 23]{cm:Schneier1996cm}). 
      Alice and Bob perform all 
      2-person-protocols. Alice will initiate all protocols and 
      Bob answers. The attackers are named Eve (eavesdropper) and
      Mallory (malicious active attacker).
  } wants to communicate with Bob, then she finds Bob's public key 
in the directory and uses it to encrypt her message to him. She then sends
this cipher text to Bob, who is then able to decrypt it again using his 
secret key. As only Bob knows his secret key, only he can decrypt 
messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

@@

\bibitem[Biryukov2009]{cm:Biryukov2009} \index{Biryukov 2009}
	Alex Biryukov, Dmitry Khovratovich, \\
	{\em Related-key Cryptanalysis of the Full AES-192 and AES-256}, \\
	2009 \\
	\url{http://eprint.iacr.org/2009/317}

\bibitem[Coppersmith2002]{cm:Coppersmith2002}  \index{Coppersmith 2002}
       Don Coppersmith, \\
       {\em Re: Impact of Courtois and Pieprzyk results}, \\
       2002-09-19, ``AES Discussion Groups''~ at \\
       \href{http://aes.nist.gov/aes/}
       {\texttt{http://aes.nist.gov/aes/}}

\bibitem[Courtois2002]{cm:Courtois2002}  \index{Courtois 2002}
        Nicolas Courtois, Josef Pieprzyk, \\
        {\em Cryptanalysis of Block Ciphers with Overdefined Systems
             of Equations}, \\
        received 10 Apr 2002, last revised 9 Nov 2002.\\
        A different version, so called compact version of the first XSL attack,
       was published at Asiacrypt Dec 2002. \\
       \href{http://eprint.iacr.org/2002/044}
       {\texttt{http://eprint.iacr.org/2002/044}}

\bibitem[Ferguson2001]{cm:Ferguson2001}  \index{Ferguson 2001}
       Niels Ferguson, Richard Schroeppel, Doug Whiting, \\
       {\em A simple algebraic representation of Rijndael}, 
       Draft 2001/05/1, \\
       \href{http://www.xs4all.nl/~vorpal/pubs/rdalgeq.html}
       {\texttt{http://www.xs4all.nl/\~{}vorpal/pubs/rdalgeq.html}}

\bibitem[Lucks-DuD2002]{cm:Lucks-DuD2002}  \index{Lucks 2002}
       Stefan Lucks, R\"udiger Weis, \\
       {\em Neue Ergebnisse zur Sicherheit des Verschl\"usselungsstandards AES}, 
       in DuD Dec. 2002.

\bibitem[Nichols1996]{cm:Nichols1996} \index{Nichols 1996} 
       Randall K. Nichols, \\
       {\em Classical Cryptography Course, Volume 1 and 2}, \\
       Aegean Park Press 1996;
       or in 12 lessons online at \\
       \href{http://www.fortunecity.com/skyscraper/coding/379/lesson1.htm}
       {\texttt{http://www.fortunecity.com/skyscraper/coding/379/lesson1.htm}}

\bibitem[Robshaw2002a]{cm:Robshaw2002a}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Essential Algebraic Structure within the AES}, 
       June 5, 2002, Crypto 2002,  \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

\bibitem[Robshaw2002b]{cm:Robshaw2002b}  \index{Robshaw 2002}
       S.P. Murphy, M.J.B. Robshaw, \\
       {\em Comments on the Security of the AES and the XSL Technique}, 
       September 26, 2002, \\
       \href{http://www.isg.rhul.ac.uk/\~{}mrobshaw/rijndael/rijndael.html}
       {\texttt{http://www.isg.rhul.ac.uk/~mrobshaw/rijndael/rijndael.html}}

\bibitem[Schmeh2003]{cm:Schmeh2003}  \index{Schmeh 2003}
       Klaus Schmeh, \\
       {\em Cryptography and Public Key Infrastructures on the Internet},\\ 
       John Wiley \& Sons Ltd., Chichester 2003. \\
       An up-to-date, easy to read book, which also considers practical
       problems such as standardisation or real existing software.

\bibitem[Schneier1996]{cm:Schneier1996cm} \index{Schneier 1996} 
       Bruce Schneier, \\
       {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
       Wiley 1994, 2nd edition 1996.

\bibitem[Stamp2007]{cm:Stamp2007} \index{Stamp 2007} 
       Mark Stamp, Richard M. Low, \\
       {\em Applied Cryptanalysis: Breaking Ciphers in the Real World}, \\
       Wiley-IEEE Press 2007, \\
       \href{http://cs.sjsu.edu/faculty/stamp/crypto/}
       {\texttt{http://cs.sjsu.edu/faculty/stamp/crypto/}}

\bibitem[Swenson2008]{cm:Swenson2008} \index{Swenson 2008} 
       Christopher Swenson, \\
       {\em Modern Cryptanalysis: Techniques for Advanced Code Breaking}, \\
       Wiley 2008.

\bibitem[Wobst-iX2002]{cm:Wobst-iX2002}  \index{Wobst 2002}
       Reinhard Wobst, \\
       {\em Angekratzt - Kryptoanalyse von AES schreitet voran}, 
       in iX Dec. 2002, \\
       plus the reader's remark by Johannes Merkle in iX Feb. 2003.

### 2009-08-06 script/en/chapters/digitalsignatures.tex
resolved more pdfTeX warnings by adding a chapter specific prefix "xy:" to all cite and bibitem in chapters with bibliography

During the past several years no new attacks against hash algorithms
came up, and so the candidates that had not yet shown any weaknesses
in their structure in practice 
(e.g. \index{SHA-1} SHA--1\footnote{%
  SHA-1 \index{SHA-1} is a 160 bit hash function specified in FIPS 180-1
  (by NIST), ANSI X9.30 Part 2 and \cite{FIPS186}.\\
  SHA means Secure Hash Algorithm, and is widely used, 
  e.g. with DSA, RSA or ECDSA.\\
  The current standard \cite{FIPS180-3} defines four secure hash algorithms
  -- SHA-1, SHA-256, SHA-384, and SHA-512.
  For these hash algorithms there are also validation tests defined in the
  test suite FIPS 140-2.

@@

The initially published result reduced the expected complexity for one SHA-1
collision search from $2^{80}$ (brute force) to $2^{69}$
\cite{Wang2005}. More recent announcements claim to futher reduce the required
effort to $2^{63}$ \cite{Wang2005b} and $2^{52}$ \cite{McDonald2009}. This would
bring collision attacks into the practial real, as similar efforts have been
mastered in the past (s.\ \ref{Brute-force-gegen-Symmetr}).

@@

Already before Crypto 2004 the U.S. National Institute of Standards and
Technology (NIST)\index{NIST} announced, to discontinue SHA-1 in the next few
years. So it is recommended not to use SHA-1 for new products generating digital
signatures. The SHA-2 family \cite{FIPS180-3} provides stronger algorithms. To
address new findings in crytanalysis, NIST has opened a competition to develop a
new cryptographic hash algorithm ``SHA-3'' in 2008. The competition is expected
to end in 2012.\footnote{\url{http://csrc.nist.gov/groups/ST/hash/sha-3/}}

@@

As explained above, signatures are not performed on the message itself, but
on a cryptographic hash value of the message. To prevent certain attacks
on the signature procedure (alone or in combination with encryption) it is
necessary to format the hash value before doing the exponentiation, as
described in the PKCS\#1 (Public Key Cryptography Standard \#1
\cite{PKCS1})\index{PKCS\#1}. 
The fact that this standard had to be revised recently, after
being in use for several years, can serve as an example of how difficult it is
to get the details of cryptography right.

@@

In August of 1991, the U.S. National Institute of Standards and Technology
(NIST\index{NIST}) proposed a digital signature algorithm (DSA), which was
subsequently adopted as a U.S. Federal Information Processing Standard (FIPS 186
\cite{FIPS186}).

@@

While DSA was specifically designed, so that it can be exported from
countries regulating export of encryption soft and hardware (like the U.S.\ 
at the time when it was specified), it has been noted
\cite[p.~490]{Schneier1996ds}, that the operations involved in DSA can be
used to emulate RSA and ElGamal encryption.

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\subsection{X.509 certificate}
\index{X.509} 
Each participant who wants to have an X.509 certificate (\cite{X.509})
verifying that his public key belongs to a real person consults what is
known as a \index{Certification authority (CA)} certification authority
(CA)\footnote{%
Often called trust center, if the certificates are not only offered
to a closed user group.
}. 
He proves his identity to this CA (for example by showing his ID). 
The CA then issues him an electronic document (certificate) which 
essentially contains the name of the certificate-holder and the name
of the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

@@

\bibitem[FIPS180-3]{FIPS180-3} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS180-3} \\
    {\em Secure Hash Standard (SHS)}, \\
    October 2008.\\
    (FIPS 180-3 supersedes FIPS 180-2.)

\bibitem[FIPS186]{FIPS186} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS186} \\
    {\em Entity authentication using public key cryptography}, \\
    February 18, 1997.\\
    No more valid.

\bibitem[FIPS186-2]{FIPS186-2} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS186-2} \\
    {\em Digital Signature Standard (DSS)}, \\
    January 27, 2000. Change Note: October 5, 2001.\\
  \href{http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf}
   {\tt http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf}

\bibitem[McDonald2009]{McDonald2009} Cameron McDonald, Philip Hawkes, Josef Pieprzyk, 
    \index{McDonald 2009} \\
    {\em Differential Path for SHA-1 with complexity $O(2^{52})$}, \\
    \url{http://eprint.iacr.org/2009/259}

\bibitem[PKCS1]{PKCS1} RSA Laboratories,
    \index{PKCS\#1} \index{RSA Laboratories} \\
    {\em PKCS \#1 v2.1 Draft 3: RSA Cryptography Standard}, \\
    April 19, 2002.

\bibitem[Schneier1996]{Schneier1996ds} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley, 2nd edition, 1996.

\bibitem[Wang2005]{Wang2005} Xiaoyun Wang, Yiqun Yin, Hongbo Yu, 
    \index{Wang 2005} \\
    {\em Finding Collisions in the Full SHA-1}, \\
    Advances in Cryptology-Crypto 2005, LNCS 3621: 17-36, 2005.

\bibitem[Wang2005b]{Wang2005b}  Xiaoyun Wang, Andrew Yao and Frances Yao, 
    \index{Wang 2005} \\
    {\em New Collision Search for SHA-1}, \\
    Crypto 2005 Rump Session \\
    \url{http://www.iacr.org/conferences/crypto2005/rumpSchedule.html}

\bibitem[Wobst2005]{Wobst2005} \index{Wobst 2005} 
    Reinhard Wobst, \\
    {\em New Attacks Against Hash Functions}, \\
    Information Security Bulletin, April 2005.

\bibitem[X.509]{X.509} ITU-T, 
    \index{X.509} \\
    {\em ITU-T Recommendation X.509 (1997 E): Information Technology -- 
    Open Systems Interconnection -- The Directory: Authentication Framework},\\
    June 1997.

\bibitem[X.509v3]{X.509v3} ITU-T, 
    \index{X.509} \index{ITU-T} \index{ISO/IEC 9594-8} \\
    {\em X.509 (1993) Amendment 1: Certificate Extensions, The Directory
    Authentication Framework},\\ 
    International Telecommunication Union, Geneva, Switzerland, July 1995\\
    (equivalent to amendment 1 to ISO/IEC 9594-8).

>>>>

During the past several years no new attacks against hash algorithms
came up, and so the candidates that had not yet shown any weaknesses
in their structure in practice 
(e.g. \index{SHA-1} SHA--1\footnote{%
  SHA-1 \index{SHA-1} is a 160 bit hash function specified in FIPS 180-1
  (by NIST), ANSI X9.30 Part 2 and \cite{ds:FIPS186}.\\
  SHA means Secure Hash Algorithm, and is widely used, 
  e.g. with DSA, RSA or ECDSA.\\
  The current standard \cite{ds:FIPS180-3} defines four secure hash algorithms
  -- SHA-1, SHA-256, SHA-384, and SHA-512.
  For these hash algorithms there are also validation tests defined in the
  test suite FIPS 140-2.

@@

The initially published result reduced the expected complexity for one SHA-1
collision search from $2^{80}$ (brute force) to $2^{69}$
\cite{ds:Wang2005}. More recent announcements claim to futher reduce the required
effort to $2^{63}$ \cite{ds:Wang2005b} and $2^{52}$ \cite{ds:McDonald2009}. This would
bring collision attacks into the practial real, as similar efforts have been
mastered in the past (s.\ \ref{Brute-force-gegen-Symmetr}).

@@

Already before Crypto 2004 the U.S. National Institute of Standards and
Technology (NIST)\index{NIST} announced, to discontinue SHA-1 in the next few
years. So it is recommended not to use SHA-1 for new products generating digital
signatures. The SHA-2 family \cite{ds:FIPS180-3} provides stronger algorithms. To
address new findings in crytanalysis, NIST has opened a competition to develop a
new cryptographic hash algorithm ``SHA-3'' in 2008. The competition is expected
to end in 2012.\footnote{\url{http://csrc.nist.gov/groups/ST/hash/sha-3/}}

@@

As explained above, signatures are not performed on the message itself, but
on a cryptographic hash value of the message. To prevent certain attacks
on the signature procedure (alone or in combination with encryption) it is
necessary to format the hash value before doing the exponentiation, as
described in the PKCS\#1 (Public Key Cryptography Standard \#1
\cite{ds:PKCS1})\index{PKCS\#1}. 
The fact that this standard had to be revised recently, after
being in use for several years, can serve as an example of how difficult it is
to get the details of cryptography right.

@@

In August of 1991, the U.S. National Institute of Standards and Technology
(NIST\index{NIST}) proposed a digital signature algorithm (DSA), which was
subsequently adopted as a U.S. Federal Information Processing Standard (FIPS 186
\cite{ds:FIPS186}).

@@

While DSA was specifically designed, so that it can be exported from
countries regulating export of encryption soft and hardware (like the U.S.\ 
at the time when it was specified), it has been noted
\cite[p.~490]{ds:Schneier1996ds}, that the operations involved in DSA can be
used to emulate RSA and ElGamal encryption.

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\subsection{X.509 certificate}
\index{X.509} 
Each participant who wants to have an X.509 certificate (\cite{ds:X.509})
verifying that his public key belongs to a real person consults what is
known as a \index{Certification authority (CA)} certification authority
(CA)\footnote{%
Often called trust center, if the certificates are not only offered
to a closed user group.
}. 
He proves his identity to this CA (for example by showing his ID). 
The CA then issues him an electronic document (certificate) which 
essentially contains the name of the certificate-holder and the name
of the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

@@

\bibitem[FIPS180-3]{ds:FIPS180-3} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS180-3} \\
    {\em Secure Hash Standard (SHS)}, \\
    October 2008.\\
    (FIPS 180-3 supersedes FIPS 180-2.)

\bibitem[FIPS186]{ds:FIPS186} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS186} \\
    {\em Entity authentication using public key cryptography}, \\
    February 18, 1997.\\
    No more valid.

\bibitem[FIPS186-2]{ds:FIPS186-2} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS186-2} \\
    {\em Digital Signature Standard (DSS)}, \\
    January 27, 2000. Change Note: October 5, 2001.\\
  \href{http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf}
   {\tt http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf}

\bibitem[McDonald2009]{ds:McDonald2009} Cameron McDonald, Philip Hawkes, Josef Pieprzyk, 
    \index{McDonald 2009} \\
    {\em Differential Path for SHA-1 with complexity $O(2^{52})$}, \\
    \url{http://eprint.iacr.org/2009/259}

\bibitem[PKCS1]{ds:PKCS1} RSA Laboratories,
    \index{PKCS\#1} \index{RSA Laboratories} \\
    {\em PKCS \#1 v2.1 Draft 3: RSA Cryptography Standard}, \\
    April 19, 2002.

\bibitem[Schneier1996]{ds:Schneier1996ds} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley, 2nd edition, 1996.

\bibitem[Wang2005]{ds:Wang2005} Xiaoyun Wang, Yiqun Yin, Hongbo Yu, 
    \index{Wang 2005} \\
    {\em Finding Collisions in the Full SHA-1}, \\
    Advances in Cryptology-Crypto 2005, LNCS 3621: 17-36, 2005.

\bibitem[Wang2005b]{ds:Wang2005b}  Xiaoyun Wang, Andrew Yao and Frances Yao, 
    \index{Wang 2005} \\
    {\em New Collision Search for SHA-1}, \\
    Crypto 2005 Rump Session \\
    \url{http://www.iacr.org/conferences/crypto2005/rumpSchedule.html}

\bibitem[Wobst2005]{ds:Wobst2005} \index{Wobst 2005} 
    Reinhard Wobst, \\
    {\em New Attacks Against Hash Functions}, \\
    Information Security Bulletin, April 2005.

\bibitem[X.509]{ds:X.509} ITU-T, 
    \index{X.509} \\
    {\em ITU-T Recommendation X.509 (1997 E): Information Technology -- 
    Open Systems Interconnection -- The Directory: Authentication Framework},\\
    June 1997.

\bibitem[X.509v3]{ds:X.509v3} ITU-T, 
    \index{X.509} \index{ITU-T} \index{ISO/IEC 9594-8} \\
    {\em X.509 (1993) Amendment 1: Certificate Extensions, The Directory
    Authentication Framework},\\ 
    International Telecommunication Union, Geneva, Switzerland, July 1995\\
    (equivalent to amendment 1 to ISO/IEC 9594-8).
